---
title: "Danish Kings"
author: "Alexander Hilmer"
date: "13/3/2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the kings

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the

separator:The Seperator is a comma (`,`).  

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
library(tidyverse)
# FILL IN THE CODE BELOW and review the outputs
kings1 <- read.csv("data/kongerækken.csv")


kings2 <- read_csv("data/kongerækken.csv")


kings3 <- read.csv2("data/kongerækken.csv")


kings4 <- read_csv2( "data/kongerækken.csv")

```
Answer: 
1. Which of these functions is a `tidyverse` function? Read data with it below into a `kings` object

read_csv2 and read_csv are tidyverse functions. I use the read_csv function to read the data into a kings object.

2. What is the result of running `class()` on the `kings` object created with a tidyverse function.

The result of running `class()` on the `kings` object created with a tidyverse function is "spec_tbl_df", "tbl_df", "tbl", "data.frame".

3. How many columns does the object have when created with these different functions?

Using the glimpse function i can see that has 5 columns 

4. Show the dataset so that we can see how R interprets each column

You can see the dataset with glimpse or for a better view use the view function


```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

kings <- read_csv("data/kongerækken.csv",na = "NULL")

class(kings)


glimpse(kings)

```


## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r eval=TRUE}
# YOUR CODE
#My data is messy since the numbers are considered as characters. I will fix it by using the as.numeric function to convert the data to numbers.

kings$birth_year  <- as.numeric(kings$birth_year)
kings$death_year  <- as.numeric(kings$death_year)
kings$reign_start <- as.numeric(kings$reign_start)
kings$reign_end   <- as.numeric(kings$reign_end)

kings_duration <- kings %>%
  mutate(duration = reign_end - reign_start)

glimpse(kings_duration)
```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `kings` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`


```{r eval=TRUE}
# YOUR CODE

average_duration <- kings_duration %>%
  summarise(avg_reign = mean(duration, na.rm = TRUE))

glimpse(average_duration)


```

## How many and which kings enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `kings` dataset. Display the result and also  count the resulting rows with `count()`

```{r eval=TRUE}
# YOUR CODE

#Start by making a object with the average duration of reign

avg_duration <- mean(kings_duration$duration, na.rm = TRUE)

#Then filter the kings with a longer than average duration of reign

longer_than_avg <- kings_duration %>%
  filter(duration > avg_duration)

#using the count function to count the number of rows in the longer_than_avg dataset to see how many kings enjoyed a longer-than-average duration of reign
count(longer_than_avg)

#Now to see which kings enjoyed a longer-than-average duration of reign, we can select the relevant columns to display

longer_than_avg %>%
  select(name, duration)


```


## How many days did the three longest-ruling monarchs rule? 

Christian 4. ruled for 21915 days	
Margrethe 2. ruled for 18993 days
Christian 9. ruled for 15705 days

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r eval=TRUE}
# YOUR CODE

top_kings <- kings_duration %>%
  arrange(desc(duration)) 

glimpse(top_kings)

#Select the three longest-ruling monarchs and calculate the total number of days they ruled

mutate(top_kings, Days = duration * 365 + (duration %/% 4))

#(duration %/% 4)) is used to account for the leap years. 


```


## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r}
# YOUR CODE
```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!


